Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeast Taiwan

### 2019-11-19 analysis/paper/paper.Rmd
add semilandmark and outline approaches, specialization of pottery

Geometric morphometry methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests. Different from linear measurement approaches that capture shapes by measuring distances, angles or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates with their unique location and orientation with respect to numerical values of coordinate axes. According to different focuses and forms of original shape data, there are two common morphometric methods used to capture and analyse the shapes, landmark approaches and outline approaches [@Adams2004]. Landmark approaches assign a set of landmarks onto specimens as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions or x, y, z coordinates for three dimensions. Based on the distances and angles of landmark points, the objects are translated, rotated, and scaled into the same orientation and size using generalized Procrustes analysis (GPA) [@Bookstein1991]. GPA method extracts shape information and compares shapes by taking sets of landmark configurations into a common coordinate system, where coordinate values represent shape variables that can be further analysed by further multivariate statistical analyses [@Slice2007]. Landmark approach is a useful method for archaeological objects with extreme morphological signature or distinctive curvature that provide suitable locations for landmark placement, such as projectile point tips or biologically meaningful points indicating osteological differences [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].

However, landmark approaches have a limitation that a set of landmarks may not be able to capture the shape differences of a morphological structure when the curve between landmarks conveys more important information than landmarks themselves. In addition, suitable locations for landmark placement may not be available for some complex or rounded shaped structures, such as pottery.

- related to the issue of specialization  
- CV is hard to test, we apply some statistics testing for it

@@

The scanned pottery drawings of reconstructed pots that mostly covers from rim to bottom were acquired from Bureau of Cultural Affairs in Yilan. All drawing presents two-dimensional view of the section of a vessel with indications of metric measurements. The scanned drawings were imported into the Inkscape software for outlines tracing to remove some information such as marks, lines, and numbers on the original drawing. Each traced half cross-section image was duplicated, flipped, and then joined with another one to create a 2D closed outline that can be processed later in the R software (www.rproject.org, Core-Team, 2015). Geometric morphometric analyses were conducted using the functions included in the Momocs, a R package intended to quantify the shape and compare its variation, especially for outline analysis [@Bonhomme2014]. The digitised outlines were converted into a list of successsive x-y pixel coordinates for elliptic Fourier analysis (EFA), which assesses morphological differences among pottery shapes from three occupation contexts. EFA is a powerful method to extract geometric information where closed outlines are considered as periodic functions and decomposed into a series of trigonometric functions referred as harmonic coefficients [@Claude2008; @Bonhomme2014]. The number of harmonics determines the quality of the geometry that more harmonics can capture the shapes more precisely [@Bonhomme2013]. Here we used \13\ harmonics to describe pottery shapes.

@@

# Normalise the shapes, from 
# https://www.repository.cam.ac.uk/bitstream/handle/1810/266423/Supplementary%20Code%20S1.pdf?sequence=20&isAllowed=y
jpgs_imported_coo <- 
  jpgs_imported %>% 
   Out(., fac = phases_isect)  %>% 
  # Centring, scaling and sampling pseudo-landmarks
   coo_center %>% 
   coo_scale %>% 
   coo_sample(., 1000) %>% 
  # Procrustes superimposition
   fgProcrustes %>% 
  # Normalisation of the starting points
   coo_slidedirection

@@

op <- efourier(jpgs_imported_coo, norm = FALSE)

# The progressive capture of shape geometry along the number of harmonics can be visualized with:

calibrate_reconstructions_efourier(jpgs_imported_coo,  
                          id = 1, 
                          range = c(1, 2, 4, 8, 14, 20))

# mean shape, per group
bot.ms <- mshapes(op, 2)
bot.ms$shp$`Post-European`   <- bot.ms$shp$`Post-European` %T>% coo_plot(border="blue")
`Post-European`  <- bot.ms$shp$`Post-European`  %T>% coo_draw(border="blue")
`Pre-European` <-   bot.ms$shp$`Pre-European`  %T>% coo_draw(border="red")
legend("topright", 
       lwd=1,
       col=c("blue", "red"), 
       legend=c( "Post-European", "Pre-European"))

>>>>

Geometric morphometric methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests. Different from linear measurement approaches that capture shapes by measuring length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates with their unique location and orientation with respect to numerical values of coordinate axes. According to different focuses and nature of original shape data, there are two common morphometric methods used to capture and analyse the shapes, landmark approaches and outline approaches [@Adams2004]. Landmark approaches assign a set of landmarks onto objects as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions or x, y, z coordinates for three dimensions. Based on raw landmarks, objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks is minimized using generalized Procrustes analysis (GPA) [@Bookstein1991]. GPA superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables that can be further analysed by multivariate statistics [@Slice2007]. Landmark-based morphometrics are widely used for archaeological objects with extreme morphological signature or distinctive points that provide suitable locations for landmark placement, such as projectile point tips or biologically meaningful points indicating osteological differences [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].

However, one limitation of landmark-based morphometrics is that landmark data may not be able to capture the shape differences of a morphological structure of which the curving outlines between landmarks convey important information. In addition, obvious and reliable landmark points may be unavailable for some complex, rounded or globular shaped structures. In those cases, outline approaches are more suitable to study shape by focusing more on on the overall shape of a morphological structure. One of the outline approaches is semi-landmarks method, which assign points along a outline curve at defined intervals between two landmarks to capture the variations of the outlines [@Bookstein1997; @Lawing2010]. Another approach, also the method used in this paper, is elliptic Fourier Analysis (EFA) commonly used for two-dimensional closed shape [@Kuhl1982]. EFA is a powerful method to extract geometric information where outlines are considered as periodic functions and decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adam2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality of the geometry that more harmonics can capture the shapes more precisely. The use of EFA in archaeology are mostly for bone morphology to understand the biological variation and recently more for stone artefacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Outline approach based on EFA is more suitable for shapes lacking of obvious landmarks or curves convey more important information indicating the meaningful variation in shapes, for example, ceramics. @Wilczek2014 evaluate the concordance between outline-based approaches, including the EFA and the Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels in varied forms from the Bibracte oppidum in France. They found that the variation in ceramic shapes analysed by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic forms. EFA enables the study of the level of production standardisation to answer questions about the evolution of shape over space and time.

In this paper, we use EFA for pottery shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeast Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production. Craft specialization can provide further evidence about the type of production organization which is closely related to economic or political aspects of a society [@Arnold2000; @Blackman1993; @Costin1991; @Hirshman2010; @Stark1995]. The emergence of craft specialization can also be an indicator for social inequality that hints the possibility of the presence of production groups worked for specific individuals. The presence of craft specialization is commonly studied by measuring the standardization of ceramics based on the assumption that the specialized mass production will lead to uniformity or homogeneity of the product due to increased skills, routinization, and lower diversity of producers [@Arnold2000]. To test the prediction of a shift in the level of social inequality in this case study in northeast Taiwan influnced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistent in forms.

- CV is hard to test, we apply some statistics testing for it

@@

The scanned pottery drawings of reconstructed pots that mostly covers from rim to bottom were acquired from Bureau of Cultural Affairs in Yilan. All drawing presents two-dimensional view of the section of a vessel with indications of metric measurements. The scanned drawings were imported into the Inkscape software for outlines tracing to remove some information such as marks, lines, and numbers on the original drawing. Each traced half cross-section image was duplicated, flipped, and then joined with another one to create a 2D closed outline that can be processed later in the R software (www.rproject.org, Core-Team, 2015). Geometric morphometric analyses were conducted using the functions included in the Momocs, a R package intended to quantify the shape and compare its variation, especially for outline analysis [@Bonhomme2014]. The digitised outlines were converted into a list of successsive x-y pixel coordinates for elliptic Fourier analysis (EFA), which assesses morphological differences among pottery shapes from three occupation contexts. Here we used \13\ harmonics to describe pottery shapes.

@@

# Normalise the shapes, from 
# https://www.repository.cam.ac.uk/bitstream/handle/1810/266423/Supplementary%20Code%20S1.pdf?sequence=20&isAllowed=y
jpgs_imported_coo <- 
  jpgs_imported %>% 
   Out(., fac = phases_isect)  %>% 
  # Centring, scaling and sampling pseudo-landmarks
   coo_center %>% 
   coo_scale %>% 
   coo_sample(., 1000) %>% 
  # Procrustes superimposition for pseudo-landmarks
   fgProcrustes %>% 
  # Normalisation of the starting points
   coo_slidedirection

@@

# take a look when they stack together
stack(jpgs_imported_coo)

# fit the x and y coordinates separately
coo_oscillo(jpgs_imported_coo[1], "efourier")

# default is 13 harmanics that gather 99% of the harmonic power 
op <- efourier(jpgs_imported_coo, norm = FALSE)

# The progressive capture of shape geometry along the number of harmonics can be visualized with:
calibrate_reconstructions_efourier(jpgs_imported_coo,  
                          id = 1, 
                          range = c(1, 2, 4, 8, 14, 20))

# mean shape, per group
bot.ms <- mshapes(op, 2)
bot.ms$shp$`Post-European` <- bot.ms$shp$`Post-European`  %>% coo_plot(border="blue")
`Post-European`  <- bot.ms$shp$`Post-European`  %>% coo_draw(border="blue")
`Pre-European` <-   bot.ms$shp$`Pre-European`  %>% coo_draw(border="red")
legend("topright", 
       lwd=1,
       col=c("blue", "red"), 
       legend=c( "Post-European", "Pre-European"))

# or from mshapes directly
jpgs_imported_coo %>% efourier(6) %>% mshapes(~phase) %>% plot_mshapes()
